---
title: "Cohort2Trajectory"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{a07_Cohort2Trajectory}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

R-package [Cohort2Trajectory](https://github.com/HealthInformaticsUT/Cohort2Trajectory) creates patient orientated treatment trajectories from cohorts defined in OHDSI ATLAS. The package can be used with access to OMOP CDM database. The package creates discrete and continuous time trajectories (and outputs them as .csv file) which describe patients' treatment through time.

## Usage

We have created wrappers to use the `CohortContrast` output for creating trajectories in `Cohort2Trajectory`. NOTE! It will use only the concepts which are inside the `data_feature` dataframe and reported in `trajectoryDataList` object. The same features are shown on plots if you load the `CohortContrast` object in the graphical user interface.


Firstly, we have to initiate the database connection and load (or create) the `CohortContrast` object.

```{r, include = TRUE, eval=FALSE, echo=TRUE}
# Initiate database connection & load data
pathToResults <- "./inst/example"
db <- DBI::dbConnect(duckdb::duckdb(), dbdir = CDMConnector::eunomia_dir("GiBleed"))
cdm <- CDMConnector::cdm_from_con(db , cdm_name = "eunomia", cdm_schema = "main", write_schema = "main")
data <- readRDS("./inst/example/example.rds")
```

Secondly, we have to import the data for the trajectories from the database.

```{r, include = TRUE, eval=FALSE, echo=TRUE}
# Import raw data for trajectory creation
data <- CohortContrast::createC2TInput(data, cdm)
```

Thirdly, we will call `Cohort2Trajectory` execution function. It has many parameters, but an easy example is given here, for more info check references.

```{r, include = TRUE, eval=FALSE, echo=TRUE}
# Create trajectories
trajectories <- CohortContrast::C2TCaller(
  data,
  pathToResults
)
```

This will output a trajectory dataframe as well as save this a a csv file to the `pathToResults` path.
